---
title: "Package 'bcprocs'"
author: "Yiran Sheng"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: yes
  pdf_document:
    highlight: zenburn
    toc: yes
vignette: |
  %\VignetteIndexEntry{Vignette Title} %\VignetteEngine{knitr::knitr} %\usepackage[utf8]{inputenc}
---
```{r, setup, include=FALSE}
knitr::opts_chunk$set(cache=FALSE)
# Loading packages and Data
library(pander)
library(bcprocs)
```

****************************************

# Procedures Overview


**bcprocs.r-package** 

*Buzaianu and Chen procedures for finding the best treatment among experimental Bernoulli populations* 


**Installation Locally**

```{r, eval=FALSE}
install.packages()
```

#### Problem Formalization

Let $\pi_1, \pi_2, \dots, \pi_k$ be k independent binomial populations with unknown parameters $\theta_1, \theta_2, \dots, \theta_n$. In addition, we have a control population $\pi_0$ with parameter $\theta_0$. The goal is to selection the best treatment with "superior" probability of success compared to $\pi_0$ or control provided such a treatment exists.  

#### Assumptions for the Experiment

Following the framework laid out by Thall et al. (1988) and Buzaianu and Chen (2008), the assumptions for the experimental procedures are:

* The experimenter should specify constants $\delta_0, \delta_1, \delta_2$, with $0 < \delta_0 < \delta_1$, the value $\theta_0 + \delta_1$ indicates only a *marginal* improvement over control, while $\theta_0 + \delta_2$ represents a significant improvement over control 

* The experimenter should also specify constants $P^{*}_1$ and $P^{*}_2$ $((2^{-k}) < P^*_1 <1, 1-2^{-k}<P^*_2<1)$, so that $\Pr\{\text{Select Control}\} \geq P^*_1$ if $max_{i \in K}(\theta_i) \leq \theta_0+\delta_0$ and $\Pr\{\text{Select} \pi_k\} \geq P^*_2$ if $max_{i \in K}(\theta_k) \geq \theta_0+\delta_2$

* No $\theta_i$ lies in the interval $(\theta_0+\delta_1, \theta_0+\delta_2)$

In the context of hypothesis testing $P^*_1$ is equivalent to $1-\alpha$ where $\alpha$ is the size of the test, and $P^*_2$ is equivalent to the power of the test $1-\beta$. 

# Proposed Multistage Procedure






```{r, results='asis', echo=FALSE}
procs.info <- data.frame( rbind(
  c('$R_{bc1}$', '`Rbc1`', 'Single-stage procedure without curtailment'),
  c('$R_{bc2}$', '`Rbc2`', 'Single-stage procedure with curtailment'),
  c('$R_{bc1}^{h}$', '`Rhbc1`', 'Two-stage procedure without curtailment'),
  c('$R_{bc2}^{h}$', '`Rhbc2`', 'Two-stage procedure with curtailment'),
  c('$R_{bc2}^{M}$', '`RMbc2`', 'Multistage procedure with curtailment (testing)')
))
names(procs.info) <- c('Procedure', 'S3 R Object Class', 'Description')
set.caption('List of packaged procedures') 
pander(procs.info)
```




## Procedure $R_{bc1}$

**Description**

`Rbc1` - S3 Class constructor for $R_{bc1}$ procedure.

**Arguments**


+-----------+------------------------------------------------------------------------+
| Parameter | Description                                                            |
+===========+========================================================================+
| `k`       | number of treatments                                                   |
+-----------+------------------------------------------------------------------------+
| `n`       | number of maximum observations to take from each treatment can be:     |
|           |                                                                        |
|           | - either a single positive integer                                     |
|           | - or a integer vector of length `k+1`                                  |
|           |                                                                        |
+-----------+------------------------------------------------------------------------+
| `c`       | design parameter c, cutoff point for choosing a treatment over control |
+-----------+------------------------------------------------------------------------+

**Value**

S3 Class object of the same procedure type (`Rbc1`)

**Examples**

Run on a full data-set.

```{r}
# run on a full dataset
# data is a data.frame of dimention (5 ,2), with 0/1 values
data <- data.frame(
  rbind(
    c(0, 1, 1),
    c(1, 1, 0),
    c(0, 0, 1),
    c(0, 1, 0),
    c(0, 1, 0)
  )
)
proc <- Rbc1(k=2, n=5, c=1, data=data)
summary(proc)
```

Run with vector-at-a-time sampling.

```{r}
# Simulation Example, refer to Buzaianu and Chen(2008)
# Choose: k=3, P1=0.95, P2=0.95, theta_0 = 0.2, delta_1 = 0.05, delta_2 = 0.2
# From Table 2: we have, n = 128, c=13
n <- 128
k <- 3
proc <- Rbc1(k=k, c=13, n=n)
# generate population parameters that satisfies model assumptions
repeat {
  p <- c(0.2, runif(k))         # population bernoulie parameters
  jmax <- which.max(p[-1]) + 1  # best treatment
  if (p[jmax] >= p[1] + 0.2) {  # ensure assumptions in Section (2) 
                                # are satisfied AND there's a best treatment to select
    invalid <- sapply(p[-1], function(x) {
      x > p[1] + 0.05 & x < p[1] + 0.2
    })
    if(all(!invalid)) break
  }
}
step <- 1
repeat {
  sampleObs <- sapply(p, function(prob) { 
  # generate B(1, p) binary r.v. as observations, from population probs.
    rbinom(1, 1, prob)
  })
  proc <- vector.at.a.time(proc, sampleObs)
  step <- step + 1
  if (step > n | ( is.done(proc) )) break
}
summary(proc)
```


## Procedure $R_{bc2}$

**Description**

`Rbc2` - S3 Class constructor for $R_{bc2}$ procedure.

**Arguments**


+-----------+------------------------------------------------------------------------+
| Parameter | Description                                                            |
+===========+========================================================================+
| `k`       | number of treatments                                                   |
+-----------+------------------------------------------------------------------------+
| `n`       | number of maximum observations to take from each treatment             |
+-----------+------------------------------------------------------------------------+
| `c`       | design parameter c, cutoff point for choosing a treatment over control |
+-----------+------------------------------------------------------------------------+

**Value**

S3 Class object of the same procedure type (`Rbc2`)

**Examples**

```{r}
# Simulation Example
# Choose: k=3, P1=0.95, P2=0.95, theta_0 = 0.2, delta_1 = 0.05, delta_2 = 0.2
# From Table 2: we have, n = 128, c=13
n <- 128
k <- 3
proc <- Rbc2(k=k, c=13, n=n)
repeat {
  p <- c(0.2, runif(k))         # population bernoulie parameters
  jmax <- which.max(p[-1]) + 1  # best treatment
  if (p[jmax] >= p[1] + 0.2) {  
    invalid <- sapply(p[-1], function(x) {
      x > p[1] + 0.05 & x < p[1] + 0.2
    })
    if(all(!invalid)) break
  }
}
step <- 1
repeat {
  sampleObs <- sapply(p, function(prob) { 
    rbinom(1, 1, prob)
  })
  proc <- vector.at.a.time(proc, sampleObs)
  step <- step + 1
  if (step > n | ( is.done(proc) )) break
}
summary(proc)
```

Reproducing example from See Buzaianu and Chen (2008) Section (6). 

```{r}
# See Buzaianu and Chen (2008) Section (6)
# Example using data from Table 3. 
data(rbc2.example)
proc <- Rbc2(k=4, n=112, c=13)

step <- 1
repeat {
  proc <- vector.at.a.time(proc, rbc2.example[step, ])
  step <- step + 1
  if (step > nrow(rbc2.example) || is.done(proc) ) break
}

summary(proc)
```

## Helper functions for $R_{bc1}$

We also provide two helper functions `Size` and `Power`, to compute $P_1^*$ and $P_2^*$ based on experiment design. 

**Senario 1:** $max_{i \in K}(\theta_i) \leq \theta_0+\delta_0$

`Size`$(k, c, n, \delta_0)$ := $1-P^*_1$ under LFC (least favorable condition)

LFC: $\theta_1 = \theta_2 = \dots = \theta_k = \theta_0 + \delta_0$

$P^*_1 = \Pr\{\text{Select Control}\ | \text{LFC} \}$ 

**Example**

```{r}
# Configuration from Table 2. in See Buzaianu and Chen (2008)
# ~ 0.05
Size(0.2, k=4, c=10, n=72, d0=0)
```

**Senario 2:** $max_{i \in K}(\theta_i) \geq \theta_0+\delta_2$

`Power`$(k, c, n, \delta_0, \delta_1, \delta_2)$ := $P^*_2$ under LFC (least favorable condition)

LFC: $\theta_1 = \theta_2 = \dots = \theta_{k-1} = \theta_0 + \delta_1$ and $\theta_k = \theta_0 + \delta_2$

$P^*_2 = \Pr\{\text{Correct Selection}\ | \text{LFC} \}$ 

**Example**

```{r}
# Configuration from Table 2. in See Buzaianu and Chen (2008)
# ~ 0.80
Power(0.2, k=4, c=11, n=83, d0=0, d1=0.05, d2=0.20)
```

## Procedure $R^h_{bc1}$

**Description**

`Rhbc1` - S3 Class constructor for $R_{bc1}^h$ procedure, two-stage testing and selection procedure without curtailment.

**Arguments**

+-----------+------------------------------------------------------------------------+
| Parameter | Description                                                            |
+===========+========================================================================+
| `k`       | number of treatments                                                   |
+-----------+------------------------------------------------------------------------+
| `n1`      | number of maximum observations to take from each treatment at stage 1  |
+-----------+------------------------------------------------------------------------+
| `n2`      | number of maximum observations to take from each treatment at stage 2  |
+-----------+------------------------------------------------------------------------+
| `c1`      | design parameter c1, cutoff point for choosing a treatment over control|
|           | at stage 1                                                             |
+-----------+------------------------------------------------------------------------+
| `c2`      | design parameter c2, cutoff point for choosing a treatment over control|
|           | at stage 2                                                             |
+-----------+------------------------------------------------------------------------+

**Value**

S3 Class object of the same procedure type (`Rhbc1`)

**Examples**

```{r}
# Simulated Example
k <- 3 ; n1=38 ; n2=59
repeat {
  p <- c(0.2, runif(k))         # population bernoulie parameters
  jmax <- which.max(p[-1]) + 1  # best treatment
  if (p[jmax] >= p[1] + 0.2) {  
    invalid <- sapply(p[-1], function(x) {
      x > p[1] + 0.05 & x < p[1] + 0.2
    })
    if(all(!invalid)) break
  }
}
# generate random sample for stage I
data.stage1 <- as.data.frame( sapply(p, function(p) {
  rbinom(n1, 1, p)
}) )
proc <- Rhbc1(k=k, c1=3, c2=10, n1=n1, n2=n2, data=data.stage1)
# generate random sample for stage II
data.stage2 <- as.data.frame( sapply(c(p[1], p[proc$candidate+1]), function(p) {
  rbinom(n2, 1, p)
}) )
proc <- run.experiment(proc, data.stage2)
summary(proc)
```

## Procedure $R^h_{bc2}$

**Description**

`Rhbc2` - S3 Class constructor for $R^h_{bc2}$, two-stage testing and selection procedure with curtailment.

**Arguments**


+-----------+------------------------------------------------------------------------+
| Parameter | Description                                                            |
+===========+========================================================================+
| `k`       | number of treatments                                                   |
+-----------+------------------------------------------------------------------------+
| `n1`      | number of maximum observations to take from each treatment at stage 1  |
+-----------+------------------------------------------------------------------------+
| `n2`      | number of maximum observations to take from each treatment at stage 2  |
+-----------+------------------------------------------------------------------------+
| `c1`      | design parameter c1, cutoff point for choosing a treatment over control|
|           | at stage 1                                                             |
+-----------+------------------------------------------------------------------------+
| `c2`      | design parameter c2, cutoff point for choosing a treatment over control|
|           | at stage 2                                                             |
+-----------+------------------------------------------------------------------------+

**Value**

S3 Class object of the same procedure type (`Rhbc2`)

**Examples**

Reproduce Section (6) in Buzaianu and Chen (2009)

```{r}
data(rhbc2.example)
proc <- Rhbc2(k=3, n1=52, c1=2, n2=75, c2=15)
step <- 1
repeat {
  proc <- stage.one(proc, rhbc2.example[step, ])
  step <- step + 1
  if (step > nrow(rhbc2.example) || is.done(proc, stage=1) ) break
}
# According to Buzaianu and Chen (2009)'s example
# in stage II we observe 26 successes from control
# and 64 successes from Treatment 1
x02 <- sample(c(rep(1, 26), rep(0, 64)), 90)
xk2 <- sample(c(rep(1, 64), rep(0, 21)), 85)
proc <- stage.two(proc, x02, xk2)
summary(proc)
```

Simulated Example. 

```{r}
# Simulation Example
# Choose: k=3, P1=0.95, P2=0.80, theta_0 = 0.2, delta_1 = 0.05, delta_2 = 0.2
# From Table 2: we have, n1=52, n2=75, c1=2, c2=15
k <- 3
proc <- Rhbc2(k=k, n1=52, n2=75, c1=2, c2=15)
repeat {
  p <- c(0.2, runif(k))         # population bernoulie parameters
  jmax <- which.max(p[-1]) + 1  # best treatment
  if (p[jmax] >= p[1] + 0.2) {  
    invalid <- sapply(p[-1], function(x) {
      x > p[1] + 0.05 & x < p[1] + 0.2
    })
    if(all(!invalid)) break
  }
}
step <- 1
repeat {
  sampleObs <- sapply(p, function(prob) { 
    rbinom(1, 1, prob)
  })
  proc <- stage.one(proc, sampleObs)
  step <- step + 1
  if (step > 52 || ( is.done(proc) )) break
}
summary(proc)
j <- proc$candidate + 1
n2.control <- 52+75 - proc$nobs[1]
n2.candidate <- 52+75 - proc$nobs[j] 
proc <- stage.two(proc, rbinom(n2.control, 1, p[1]), rbinom(n2.candidate, 1, p[j]))
summary(proc)
```

## Helper functions for $R^h_{bc1}$

We also provide two helper functions `Sizeh` and `Powerh`, to compute $P_1^*$ and $P_2^*$ based on experiment design. 

**Senario 1:** $max_{i \in K}(\theta_i) \leq \theta_0+\delta_0$

`Sizeh`$(k, c1, n1, c2, n2, \delta_0)$ := $1-P^*_1$ under LFC (least favorable condition)

LFC: $\theta_1 = \theta_2 = \dots = \theta_k = \theta_0 + \delta_0$

$P^*_1 = \Pr\{\text{Select Control}\ | \text{LFC} \}$ 

**Example**

```{r}
# Configuration from Table 2. in See Buzaianu and Chen (2009)
# ~ 0.05
Sizeh(0.4, k=2, c1=2, n1=47, c2=13, n2=62, d0=0)
```

**Senario 2:** $max_{i \in K}(\theta_i) \geq \theta_0+\delta_2$

`Powerh`$(k, c1, n1, c2, n2, \delta_0, \delta_1, \delta_2)$ := $P^*_2$ under LFC (least favorable condition)

LFC: $\theta_1 = \theta_2 = \dots = \theta_{k-1} = \theta_0 + \delta_1$ and $\theta_k = \theta_0 + \delta_2$

$P^*_2 = \Pr\{\text{Correct Selection}\ | \text{LFC} \}$ 

**Example**

```{r}
# Configuration from Table 2. in See Buzaianu and Chen (2009)
# ~ 0.80
Powerh(0.4, k=2, c1=2, n1=47, c2=13, n2=62, d0=0, d1=0.05, d2=0.2)
```

## Procedure $R^M_{bc2} (Experimental)$

**Description**

`RMbc2` - S3 Class constructor for $R^M_{bc2}$ procedure.

**Arguments**


+-----------+------------------------------------------------------------------------+
| Parameter | Description                                                            |
+===========+========================================================================+
| `k`       | number of treatments                                                   |
+-----------+------------------------------------------------------------------------+
| `ns`      | number of observations to take from each treatment at each stage       |
+-----------+------------------------------------------------------------------------+
| `s`       | number of maximum stages, the experiment is allowed to carry out       |
+-----------+------------------------------------------------------------------------+
| `c`       | design parameter c, cutoff point for choosing a treatment over control |
+-----------+------------------------------------------------------------------------+

**Value**

Object

- attribute `$get.state`: function that produces the current procedure state
- attribute `$next.state`: function to update procedure state, takes a single vector of binomial obs count as input
- attribute `$is.done`: function that returns Boolean value, indicating if the procedure has terminated of not

**Note**

Currently, `RMbc2` does not have `summary` generic implemented, examine procedure state manually by calling `$get.state()`

**Examples**

```{r, warning=FALSE}
p <- c(0.2, 0.17, 0.42)
proc <- RMbc2(k=2, c=8, ns=6, s=10)
repeat {
  data <- sapply(p, function(p) {
    rbinom(1, 6, p)
  })
  # eg. data -> c(2, 0, 4)
  proc$next.state(data)
  if(proc$is.done()) {
    break
  }
}
proc$get.state()$decision
```

# Refrence

1. Dunnett, C. W. (1984). Selection of the best treatment in comparison to a control with an application to a medical trial. In: Santner, T. J., Tamhane, A. C., eds. Design of Experiments: Ranking and Selection. New York: Dekker.

2. Buzaianu, E. M., Chen, P. (2005). On selecting among treatments with binomial outcomes. Commun. Statist. Theor. Meth. 34:1247–1264.

3. Buzaianu, Elena M. and Chen, Pinyuen (2008) Curtailment Procedure for Selecting Among Bernoulli Populations. Commun. Statist. Theor. Meth. 37:7, 1085 - 1102

4. Buzaianu, Elena M. and Chen, Pinyuen(2009) A Hybrid Selection and Testing Procedure with Curtailment for Comparative Clinical Trials. Sequential Analysis,28:1,2 — 20
